---
title: "Bayesian Forecasting of CO2 Emissions"
author: "Mallory Wang, Noah Kochanski"
date: "4/16/2022"
output: 
    beamer_presentation:
      theme: "AnnArbor"
      colortheme: "wolverine"
      fonttheme: "serif"
      toc: true    
      slide_level: 3
      number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::read_chunk("2_eda_vis.R")
```

```{r, Library and theme, echo = FALSE}
```
# Motivation

- Climate affects everyone on Earth
- Considered by many to be one of the biggest threats to humanity
- It becomes increasingly important to understand, track and predict our effect on global warming

# Data Introduction

### Data Introduction

- Data comes from the World Bank on the United States from 1960 to 2018
- Includes CO2 emissions, economic data (GDP, GDI, primary income), and energy consumption
- Many highly correlated variables

### Data Introduction 

- Year: annual
- C02 Emissions: measured in kt
- Population: total population
- Population in Urban Agglomeration: population in urban agglomerations of more than one million
- Economic Data:
  - Gross Domestic Product, Gross Domestic Income, Net Primary Income
- Energy Data: 
  - Energy Use, Net Energy Import, Electric Power Consumption



### Data Introduction 

```{r, facet, echo = FALSE, out.width = "70%", fig.align='center'}
```


### Data Introduction

```{r, correlation plot, echo = FALSE, fig.align='center', out.width="80%"}
```


# Frequentist Approach



# Bayesian Approach


# Comparison

